Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-07 MS/ms.Rmd
Did some small wording changes to make it flow a little better in teh new section

Nuisance heterogeneity can get in the way of understanding real biological patterns that interest comparative physiologists. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above.

@@

All searches were conducted in Scopus between May 16 and 18, 2021, with a follow up serach on September 1, 2021 (see the Supplementary Information). Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

@@

To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (*lnRR*), Standardised Mean Differences (*SMD*; e.g., Hedges *d*, *g* or Cohen's *d*), Correlation or Fisher's Z -transformed correlation (*Zr*), Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we labelled as 'raw'). All other effects were labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision about whether the effect measure was likely to be impacted. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured. The magnitude of differences between two treatments (e.g, contrast-based effect sizes -- *SMD* or *lnRR*) might depend on the pH or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected 'nuisance heterogeneity' to be common to most effect sizes. We categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random-effects meta-analysis (i.e., intercept-only model, but has random study effects); 3) multilevel meta-analysis (i.e., intercept-only model, but has multiple random effect levels such as study, species and observation-level random effects); 4) meta-regression models (i.e., a multilevel or random-effects model with fixed effects - including random effects); 5) weighted regression models (including mixed- effects models); and 6) linear regression models (including linear and generalised linear mixed-effect models, and phylogenetic least squares models) [see @NakagawaSantos2012 for discussion of models]. Notably, only model types 1--5 are weighted models because they weight effect sizes based on the inverse of their sampling variance or sample size.

>>>>

Nuisance heterogeneity can get in the way of understanding real biological patterns that interest comparative physiologists. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest then this variable can be formally incorporated in statistical models. If, however, it is not the primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above. Nonetheless, there is limited discussion about the ways in which it can be dealt with in the literature.

@@

We searched the Scopus database between May 16 and 18, 2021, with a follow up search on September 1, 2021 (see the Supplementary Information). After a series of pilot searches (detailed in Supplementary Materials) to refine our search string [See @Foo2021 for an overview of this process], we searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract. We restricted our search to key comparative physiology journals known to the authors that would likely contain literature-based quantitative syntheses within the last 6 years (detailed in Table S1 and S2 - Supplementary Materials). We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

@@

To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (*lnRR*), Standardised Mean Differences (*SMD*; e.g., Hedges *d*, *g* or Cohen's *d*), Correlation or Fisher's Z -transformed correlation (*Zr*), Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we labelled as 'raw'). All other effects were labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision about whether the effect measure was likely to be impacted. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured. The magnitude of differences between two treatments (e.g, contrast-based effect sizes -- *SMD* or *lnRR*) will depend on the pH or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected nuisance heterogeneity to be common to most effect sizes.

We also categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random-effects meta-analysis (i.e., intercept-only model, but has random study effects); 3) multilevel meta-analysis (i.e., intercept-only model, but has multiple random effect levels such as study, species and observation-level random effects); 4) meta-regression models (i.e., a multilevel or random-effects model with fixed effects - including random effects); 5) weighted regression models (including mixed- effects models); and 6) linear regression models (including linear and generalised linear mixed-effect models, and phylogenetic least squares models) [see @NakagawaSantos2012 for discussion of models]. Notably, only model types 1--5 are weighted models because they weight effect sizes based on the inverse of their sampling variance or sample size.

We summarise the results of our literature survey below while discussing solutions for dealing with nuisance heterogeneity.
